Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

### 2023-08-21 manuscript/manuscript.qmd
manuscript update
Updated Manuscript

title: Open-Source machine learning BANTER acoustic classification of a cryptic echolocating species

@@

  - name: Annamaria DeAngelis

@@

  - name: Jenny Trickey

@@

  Passive acoustic monitoring is increasingly used for assessing populations of marine mammals; however, analysis of large datasets is limited by our ability to easily classify sounds detected. Classification of beaked whale acoustic events, in particular, require evaluation of multiple lines of evidence by expert analysts. Here we present a highly automated approach to acoustic detection and classification using supervised machine learning and open source software methods. Data from four large scale surveys of beaked whales (North Atlantic, South Atlantic, Hawaii, and Eastern Pacific) were analyzed PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier). Overall correct classification results ranged from 87% for the South Atlantic data to 98% for the North Atlantic. Results for many species could likely be improved with increased sample sizes, consideration of alternative automated detectors, and addition of relevant environmental features. These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for species and geographic regions.

@@

format:
  pdf: default
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Informatics
      formatting: preprint
      model: 1p
      cite-style: authoryear

@@

Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians manually identify stereotyped sounds that could be reliably attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008]. The dramatic increase in recordings make it impossible for experienced acousticians to manually annotate all data, and there has been an increasing need for the development of automated classification routines that can provide accurate determinations of the species responsible for the sounds.

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018]. However, beaked whales make stereotyped echolocation clicks that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_spatio-temporal_2014]. Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020]. Unfortunately, manual classification of beaked whale echolocation clicks requires analysis by trained observers using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serve to improve the efficiency and decrease the cost of analyzing large datasets.

BANTER is a supervised machine learning acoustic event classifier originally developed for classifying dolphin schools from acoustic recordings [@rankin_acoustic_2017]. This hierarchical random forest event classifier relies on an initial call classifier for each call type (in the dolphin case study, this included one each for whistles, echolocation clicks, and burst pulses) followed by an event classifier. The event classifier considers the probability of assignment to species for each call in the call classifier, along with any event-level characteristics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors. While BANTER relies on multiple call types, relatively minor changes to a call detector could yield different detector results [@rankin_acoustic_2017]. In this case, settings for a whistle & moan detector were modified to improve performance at detecting burst pulses. This approach of applying multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type was analyzed), as long as the settings of the automated call detector were modified such that the results were different. This approach was successfully applied to classified echolocation clicks for narwhal and belugas [@zahn_acoustic_2021]. For species who primarily (or exclusively) produce echolocation clicks, BANTER may serve as an option for automated machine learning classification.

Here we applied BANTER acoustic classification to beaked whale detections from four large acoustic datasets including the US East Coast (north and south), Hawaii, and the US West Coast. Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER trials. Our goals are to identify an efficient and accurate automated approach to acoustic classification of beaked whales, and to provide a framework for analysis that may serve for other PAM studies.

@@

Passive acoustic data was collected during three surveys conducted by National Oceanographic and Atmospheric Administrations' research operations off the US East Coast, the Hawaiian Islands (Hawaii) and the US West Coast.

Recordings from the western North Atlantic Ocean (U.S. east coast) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey (cite). AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey area. The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (HTI-96-min, High Tech Inc., Long Beach, MS) recorded at 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see [@deangelis_description_2018] for more information). The SAtlantic area ranged from Delaware south to central Florida (GU1605, \[CITE\]) and recordings used in this analysis consisted of a single hydrophone (Reson \[MODEL\], Teledyne Marine, Slangerup, Denmark) recorded 500 kHz sample rate (soundcard) and decimated to 192 kHz with a 1 kHz high pass filter (see \[CITE\] for more information). Variation in hydrophone sensitivities precluded combining the NATlantic and SAtlantic datasets.

Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, [@yano_cetacean_2018]).  Drifting acoustic recorders were deployed at randomly selected locations within the main Hawaiian Islands. Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA). Soundtrap recorders were sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes. SM3M data were sampled at 256 kHz with continuous recordings. Individual buoys were deployed for 10 to 50 days (see [@yano_cetacean_2018] for more detailed information).

Recordings from the Eastern North Pacific Ocean off the U.S. west coast (EPacific) were collected using drifting acoustic recorders during the 2018 CCES (California Current Ecosystem Survey) survey [@simonis_passive_2020]. Buoys were deployed at 23 stations distributed throughout the California Current region; data from 15 of these included high quality acoustic data used in this analysis. Two types of drifting recorders were included: (1) Soundtrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA). Recordings were duty cycled and recorded 2 minutes out of a variable 'off' time; recorders sampled at a minimum 256 sample rate. Individual buoys were deployed for 5-79 days (see [@simonis_passive_2020] for more detailed information).

@@

All acoustic recordings were analyzed using PAMGuard software (version 2.00.15c, 2.00.16e) with a suite of generic click classifiers within the Click Detector module (see [@keating_summary_2013]).  Classifier sets were saved such that any click may be classified as more than one click type. The general click classifiers were treated as spectral band click detectors (e.g., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz). An additional detector within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see [@keating_summary_2013]).

Acoustic detection of beaked whale events were then classified to species by experienced acoustic researchers using multiple lines of evidence. If beaked whales were sighted in close proximity to the recording device, the visual confirmation of species identity was used to determine species identity. Otherwise, experienced acousticians followed strict protocol to define species identity based on acoustic characteristics. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale; only acoustic events classified to species were considered for this study. Acoustic researchers with experience in classifying beaked whale echolocation clicks examined the bearing-time display for the click detector module in PAMGuard to assess the species identity of the acoustic event. For NEFSC data, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into click trains based on identification of consecutive clicks along the same bearing angle in the bearing time plot (see [@deangelis_description_2018]). Data were saved within the PAMGuard database and binaries for downstream processing using the PAMpal package [@sakai_pampal_2021] in the R programming language [@r_core_team_r_2022].

Standard click calculations were measured from echolocation clicks using default values in PAMpal. Inter-click interval (ICI) was calculated using the 'calculateICI' function in PAMpal; the mode of the ICI was identified for each beaked whale acoustic event. In addition, for EPacific data, we used the ‘matchEnvData’ function in PAMpal to include ERDAPP environmental data (@simons_erddap_2022), including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41).

Multiple trials were conducted to better understand the value (and potential limitations) of the different suite of measurements. Data subsets for each trial were exported for BANTER classification using the export_banter function in PAMpal. Trials included standard click calculations for all clicks in a beaked whale acoustic event (EC); the addition of  ICI for each event (EC_ICI), and for the EPacific dataset we also included environmental variables (EC_ICI_ENV).

@@

BANTER models were created for each trial using the BANTER package in the R programming language [@archer_banter_2022]. Standard click calculations were used for the BANTER call classifier; event level measures of ICI and environmental variables were applied to the BANTER event classifier. A minimum of two events per species and the complete suite of variables (absence of NAs) were required for training and testing of the classification model. The BANTER random forest models contain two parameters, ntree and sampsize, which must be selected to ensure stability of the model. The sample size (sampsize) is the number of events considered for each trial of the model, and the ntree is the number of trees (iterations of the model) run in a given model. These values were modified to obtain the best results, with an emphasis on lower sample size and higher number of trees for computational performance [@rankin_banter_2021].

The model output includes the confusion matrix, and a number of visualizations provided by the rfpermute package in the R programming language [@archer_rfpermute_2022].

@@

Four study areas (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study; species included: Baird's beaked whales (*Berardius bairdii*), Blainsville's beaked whales (*Mesoplodon densirostris*), Cross Seamount beaked whale (@mcdonald_acoustic_2009), Cuvier's beaked whales (*Ziphius cavirostris*),  Gervais' beaked whales (*Mesoplodon europaeus*), Hubb's beaked whale (*Mesoplodon carlhubbsi*), Longman's beaked whale (*Indopacetus pacificus*), Sowerby's beaked whales (*Mesoplodon bidens*), Stejneger's beaked whale (*Mesoplodon stejnegeri*), True's beaked whales (*Mesoplodon mirus*), and unidentified beaked whale 'BW43' [@baumann-pickering_species-specific_2013]. Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets. Specific species combinations and sample sizes varied by study area.

@@

Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n = 120), Gervais' beaked whales (n = 4), Sowerby's beaked whales (n = 6), and True's beaked whales (n = 76). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

For the NAtlantic dataset, the model with ICI had a modest increase in accuracy over the EC (only) model (EC_ICI model, Detector Model: sampsize = 3, ntrees = 10,000; Event Model: sampsize = 3, trees = 100,000). This model provided an overall correct classification rate of 98% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a). Classification scores ranged from 75% for Gervais' beaked whales to 100% for Cuvier's and Sowerby's beaked whales. All classification results were greater than the expected error rate (Priors in @fig-natlantic a). Results from the EC_ICI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig. 1).

The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the feature space of the two 'most important' dimensions, or predictors. For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result. The importance heat map (@fig-natlantic c) shows the most important features for predicting each species. The strength of the classifications by species can be seen in the plot votes (@fig-natlantic d), which show the distribution of the votes for each of the 100,000 trials for each event. For each plot, the events are represented as vertical slices (with plot subdivided along the x-axes according to the event sample size), and the percentage of votes for each species is represented by their color along that vertical slice. In a perfect scenario, all 100,000 trees would have voted for the correct species and therefore the classification would be correct and the strength of these classifications would be maximized.

The proximity plot (@fig-natlantic b) shows overlap between Gervais', Sowerby's and True's beaked whales in this two dimension feature space and the species classifications are resolved using additional features, as shown in the importance heat map (@fig-natlantic c). The plot vote figure (@fig-natlantic d) shows strong classification strength for Cuvier's and Sowerby's beaked whales, with increased variability in voting distributions for Gervais' and True's beaked whales.

@@

Species encountered in the SAtlantic study dataset include Blainsville's beaked whales (n = 46), Cuvier's beaked whales (n = 111), Gervais' beaked whales (n = 77), Sowerby's beaked whales (n = 2), and True's beaked whales (n = 13). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

For the SAtlantic dataset, the most accurate model considered the echolocation clicks and the ICI for an event (EC_ICI Model, Detector Model: sampsize = 4, ntrees = 10,000; Event Model: sampsize = 1, ntrees = 100,000) and provided an overall correct classification rate of 87.1% for all five species. Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than the expected error rate (Priors in @fig-natlantic a). The sample sizes for the BANTER model was kept low (sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig. 3 for EC_ICI_ALT), but resulted in the loss of Sowerby's in the final model. Results from the EC_ICI model are presented here; results from the EC and EC_ICI_ALT model can be found in supplementary materials (Supplement Fig. 2).

For SAtlantic, the two most important features are insufficient to differentiate species, as seen by the overlap in species distribution with these feature spaces in the proximity plot (@fig-satlantic b). Importance variables vary by species (importance heat map @fig-satlantic c), with ICI measurements having relatively low importance for Blainsville's and Sowerby's beaked whales (@fig-satlantic c). The plot votes graph (@fig-satlantic d) shows relatively weak classification results; most events consisted of \<50% of the trees voting for the correct species.

@@

The Hawaii dataset consisted of 13 drifting buoys deployed within the main Hawaiian Islands (@yano_cetacean_2018); species included Blainsville's beaked whale (n = 521), Cross Seamount beaked whale (n = 76), Cuvier's beaked whale (n = 201), and Longman's beaked whale (n = 122). BANTER classification trials included (1) echolocation clicks (EC), and (2) echolocation clicks and ICI (EC_ICI).

The most accurate model considered the echolocation clicks only (EC Model, Detector Model: sampsize = 10, ntrees = 5,000; Event model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.3% for all four species. Classification scores ranged from a low of 86% (Cuvier's Beaked Whale) to 97.5% (Longman's beaked whale, @fig-hawaii). All classification results were greater than the expected error rate (Prior in @fig-hawaii a). Results from the EC model are presented here; results from EC_ICI model can be found in supplementary materials (Supplement Fig. 4).

Distinctive clusters based on the two most important dimensions can be found for Longman's and Blainsville's beaked whales, and there is overlap in this feature space of the clusters for Cuvier's and Cross Seamount beaked whales as shown in the Proximity Plot (@fig-hawaii b). This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the Importance Heat Map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix). The plot votes graph (@fig-hawaii d) show strong classification results for most Blainsville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.

## Eastern Pacific

The EPacific dataset consisted of 15 drifting recording buoys deployed off the west coast of the United States [@simonis_passive_2020]; species included Baird's beaked whale (n = 29), Unidentified Beaked Whale 'BW43' (n = 125), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 926), Hubbs beaked whale (n = 66) and Stejneger's beaked whale (n = 42). BANTER classification trials included (1) echolocation clicks (EC), (2) echolocation clicks and ICI (EC_ICI), and (3) echolocation clicks, ICI, and environmental features (EC_ICI_ENV).

The best model considered the echolocation clicks, ICI, and environmental features (EC_ICI_ENV Model, Detector Model: sampsize = 3, ntrees = 10,000; Event model: sampsize = 4, ntree = 10,000) and provided an overall correct classification rate of 92.2% for all six species. Classification scores ranged from a low of 91.2% (BW43) to 100% for Baird's beaked whales(@fig-epacific a). All classification results were greater than the expected error rate (Prior in @fig-epacific a). Results from the EC_ICI_ENV model are presented here; results from the EC and EC_ICI model can be found in supplementary materials (Supplement Figs. 5).

There is considerable overlap in the primary feature space for all but BW43, as shown in the Proximity Plot (@fig-epacific b). ICI was among the most important classification variable for all species (@fig-epacific c heat map). The plot votes (@fig-epacific d) show relatively strong classification results Baird's, Hubbs, and Stejneger's beaked whales.

>>>>

title: Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

@@

  - name: Annamaria I. DeAngelis

@@

  - name: Jennifer S. Trickey

@@

  Passive acoustic monitoring is increasingly used for assessing populations of marine mammals; however, analysis of large datasets is limited by our ability to easily classify sounds detected. Classification of beaked whale acoustic events, in particular, requires evaluation of multiple lines of evidence by expert analysts. Here we present a highly automated approach to acoustic detection and classification using supervised machine learning and open source software methods. Data from four large scale surveys of beaked whales (northwestern North Atlantic, southwestern North Atlantic, Hawaii, and eastern North Pacific) were analyzed using PAMGuard (acoustic detection), PAMpal (acoustic analysis) and BANTER (hierarchical random forest classifier). Overall classification accuracy ranged from 88% for the southwestern North Atlantic data to 97% for the northwestern North Atlantic. Results for many species could likely be improved with increased sample sizes, consideration of alternative automated detectors, and addition of relevant environmental features. These methods provide a highly automated approach to acoustic detection and classification using open source methods that can be readily adopted for other species and geographic regions.

@@

format: docx

#  pdf: default
#  elsevier-pdf:
#    keep-tex: true
#    journal:
#      name: Ecological Informatics
#      formatting: preprint
#      model: 1p
#      cite-style: authoryear
---

@@

Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008]. However, the dramatic increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all. This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.

Beaked whales are deep diving marine mammals found in offshore waters; their long dive intervals and cryptic surfacing behavior make them difficult to study using typical shipboard visual observation methods [@macleod_beaked_2018]. However, beaked whales make stereotyped echolocation signals in the form of frequency-modulated (FM) pulses that have temporal and spatial characteristics that can be used to differentiate species [@baumann-pickering_species-specific_2013]. Detection of beaked whale signals in large datasets is greatly expanding our understanding of their population structure and the potential impact of human activities on these species [@barlow_acoustic-based_2022]; [@baumann-pickering_spatio-temporal_2014]; [@simonis_passive_2020]. Unfortunately, manual classification of beaked whale echolocation pulses requires analysis by trained analysts using a number of visual aids to examine the call characteristics. Development of automated classification routines, if accurate, serves to improve the efficiency, reduce the subjectivity, and decrease the cost of analyzing large datasets.

BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a supervised machine learning method originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017]. This hierarchical Random Forest [@breiman_random_2001] event classifier combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter. The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors.

While BANTER was designed to classify events by integrating information from multiple types of call detectors, relatively minor changes to a detector for a specific call type could yield different detector results. In the dolphin-based model [@rankin_acoustic_2017], settings for a whistle and moan detector were modified to improve performance at detecting burst pulses. This approach of applying output from multiple call detectors (of the same type but with different settings) suggests that perhaps BANTER could be used on species that only produce one call type (or where only one call type is to be analyzed), as long as the settings of the automated call detector were modified such that the results were different. This approach was successfully applied to classified echolocation pulses for narwhals and belugas [@zahn_acoustic_2021]. For species that primarily (or exclusively) produce echolocation pulses, BANTER may serve as an option for automated machine learning classification.

Here we applied BANTER acoustic classification to beaked whale detections from large acoustic datasets from the U.S. East Coast, Hawaii, and the U.S. West Coast. Acoustic data were analyzed by experienced acousticians to determine species identity, and these 'ground truth' classifications served as training data for the supervised BANTER models. Our goals were to identify an efficient and accurate automated approach to acoustic species identification of beaked whales, and to provide a framework for analysis that may serve for other PAM studies.

@@

Passive acoustic data were collected during surveys conducted by the National Oceanographic and Atmospheric Administrations' (NOAA) research operations off the U.S. East Coast, the Hawaiian Islands, and the U.S. West Coast.

Recordings from the U.S. East Coast (western North Atlantic Ocean) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey by Northeast Fisheries Science Center and Southeast Fisheries Science Center [@united_states_national_marine_fisheries_service_2016_2016]. AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey areas. The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, Appendix A, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (HTI-96-min, High Tech Inc., Long Beach, MS) which recorded at a 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see @deangelis_description_2018 for more information). The SAtlantic area ranged from Delaware south to central Florida (GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (Reson TC4013, Teledyne Marine, Slangerup, Denmark) which recorded at a 500 kHz sample rate (custom 12 channel SailDAQ soundcard) and were decimated to 192 kHz with a 1 kHz high pass filter (see GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016] for more information). The NAtlantic and SAtlantic recordings were treated as two separate datasets due to differences in hydrophone frequency responses and sensitivities.

Recordings from Hawaii were collected using drifting acoustic recorders during the 2017 HICEAS (Hawaiian Island Cetacean Ecosystem Assessment Survey, @mccullough_acoustic_2021, @yano_cetacean_2018). Drifting acoustic recorders were deployed at randomly selected locations within the Main Hawaiian Islands. Data were collected using SoundTrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) sampled at 288 kHz with a duty cycle of 2 min on for every 10 minutes. Individual drifting recorders were deployed for 11 to 23 days (see @mccullough_acoustic_2021, @yano_cetacean_2018 for more detailed information).

Recordings from the eastern North Pacific Ocean off the U.S. West Coast (EPacific) were collected using drifting acoustic recorders during the 2018 CCES (California Current Ecosystem Survey) [@simonis_passive_2020]. Drifting recorders were deployed at 23 stations distributed throughout the California Current region; data from 15 of these included high quality acoustic data used in this analysis. Two types of drifting recorders were used: (1) SoundTrap ST4300 recorders (Ocean Acoustics, Auckland, New Zealand) and (2) Wildlife Acoustics SM3M recorders (Wildlife Acoustics, Maynard, MA). Recordings were duty cycled and recorded 2 minutes out of a variable 'off' time; recorders sampled at a minimum 256 kHz sample rate. Individual drifting recorders were deployed for 5-79 days (see @simonis_passive_2020 for more detailed information).

@@

All acoustic recordings were analyzed using PAMGuard software (versions 2.00.15c, 2.00.16e) with a suite of generic click classifiers within the Click Detector module (see @keating_summary_2013). Classifier sets were saved such that any click may be classified as more than one click type (using 'save classifier set' in Pamguard click classification window). Five general click classifiers were treated as spectral band click detectors (i.e., 2 -- 15 kHz, 15 -- 30 kHz, 30 -- 50 kHz, 50 -- 80 kHz, and \> 80 kHz). An additional sixth detector within the 30 -- 50 kHz peak frequency range considered the presence of a frequency sweep that is characteristic of beaked whale pulses (see @keating_summary_2013).

Echolocation signals that were identified as frequency-modulated pulses from beaked whales were then linked as PAMGuard "events" by experienced acoustic researchers. Events included all the pulses that were close to each other in time, had similar spectral and waveform characteristics, and were received at consistent bearing angles. For NAtlantic and SAtlantic datasets, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into putative individuals by identification of consecutive pulses along the same bearing angle in the bearing time plot (see @deangelis_using_2017).

Experienced acoustic researchers used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale. Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@r_core_team_r_2022].

Standard click metrics describing the spectral and temporal characteristics were measured from echolocation pulses using default values in *PAMpal* [@sakai_pampal_2023]. Inter-pulse interval (IPI) was calculated using the 'calculateICI' function in *PAMpal* and the mode of the IPI was identified for each beaked whale acoustic event. In addition, for the EPacific data, we used the 'matchEnvData' function in *PAMpal* to include ERDAPP environmental data [@simons_erddap_2022], including seafloor depth and seafloor gradient (both from erdSrtm30plusSeafloorGradient), as well as sea surface temperature (from jplMURSST41). A full consideration of environmental data for all regions was beyond the scope of this study.

Multiple models were created to better understand the value (and potential limitations) of the different suite of measurements. Models included standard click metrics (only) for all pulses in a beaked whale acoustic event (EC); the addition of IPI for each event (EC_IPI), and for the EPacific dataset we also included environmental variables (EC_IPI_ENV).

@@

BANTER models were created using the *banter* package [@archer_banter_2022] in R v2022.07.02 [@r_core_team_r_2022]. Output from PAMGuard click detector modules were used to create the BANTER detector models, for a total of 6 detector models (one for each click detector used). For each detector model, BANTER creates a mean of the species classification probabilities for each call type, which is applied to the BANTER event model (in conjunction with any event-level variables). A minimum of two events per species and the complete suite of variables (no missing data) were required for training and testing of the classification model. Event level variables included IPI and environmental variables (for the EPacific dataset). The Random Forest models that BANTER is based on are largely driven by two parameters, the number of trees in the forest (*ntree*) and the number of samples randomly drawn to build each tree (*sampsize*). Small values of *sampsize* were selected to improve computational performance, but large values of *ntree* were used to obtain models with stable classification results [@rankin_banter_2021]. Model results were summarized using the *rfpermute* package [@archer_rfpermute_2022].

@@

Four datasets from four different regions (NAtlantic, SAtlantic, Hawaii, EPacific) were analyzed for this study, and species included: Baird's beaked whales (*Berardius bairdii*), Blainville's beaked whales (*Mesoplodon densirostris*), Cross Seamount beaked whales [@mcdonald_acoustic_2009], Cuvier's beaked whales (*Ziphius cavirostris*), Gervais' beaked whales (*M. europaeus*), Hubbs' beaked whales (*M. carlhubbsi*), Longman's beaked whales (*Indopacetus pacificus*), Sowerby's beaked whales (*M. bidens*), Stejneger's beaked whales (*M. stejnegeri*), True's beaked whales (*M. mirus*), and unidentified beaked whale 'BW43' [@baumann-pickering_species-specific_2013]. Analyses were conducted separately for the different study areas due to differences in the data collection methods that precluded combining of datasets. Species and sample sizes varied by study area.

@@

Species encountered in the NAtlantic dataset include Cuvier's beaked whales (n = 120), Gervais' beaked whales (n = 4), Sowerby's beaked whales (n = 6), and True's beaked whales (n = 76). BANTER classification models included (1) echolocation pulses (EC), and (2) echolocation pulses and IPI (EC_IPI).

For the NAtlantic dataset, the most accurate model included echolocation pulse metrics with IPI; this model had a modest increase in accuracy over the EC (only) model (EC_IPI model, Detector Model: *sampsize* = 3, *ntree* = 10,000; Event Model: *sampsize* = 3, *ntree* = 100,000). This model provided an overall correct classification rate of 97.5% for all four species (pct.correct in Confusion Matrix, @fig-natlantic a). Classification scores ranged from 75% for Gervais' beaked whales to 100% for Cuvier's and Sowerby's beaked whales. All classification results were greater than expected (Priors in @fig-natlantic a). Results from the EC_IPI model are presented here; results from the EC model were similar and can be found in supplementary materials (Supplement Fig. 1).

The proximity plot (@fig-natlantic b) provides a view of the distribution of events within the classification model space. For each event in the plot, the color of the central dot represents the true species identity, and the circle represents the BANTER classification result. This plot shows the degree of overlap between Gervais', Sowerby's, and True's beaked whales. Species classifications of events are resolved using additional features (mean assignment probabilities for each of the detectors in the detector model, as well as any event level variables), as seen in the ten most important features for predicting each species (importance heat map, @fig-natlantic c). The strength of classifications for each event can be seen in the distribution of votes for each species across the forest (@fig-natlantic d). In each frame of this figure, the events are represented as vertical slices along the x-axis, and the percentage of votes for each species is represented by their color along that vertical slice in the y-axis. This distribution shows strong assignment probabilities for most Cuvier's and Sowerby's beaked whale events, with greater variability in assignment for Gervais' and True's beaked whale events.

@@

Species encountered in the SAtlantic study dataset include Blainville's beaked whales (n = 46), Cuvier's beaked whales (n = 111), Gervais' beaked whales (n = 77), Sowerby's beaked whales (n = 2), and True's beaked whales (n = 13). BANTER classification models included (1) echolocation pulses (EC), and (2) echolocation pulses and IPI (EC_IPI).

For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species. Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than expected (Priors in @fig-natlantic a). The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig. 3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model. Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig. 2).

For this dataset, the two most important features are insufficient to differentiate species, as seen by the overlap in species distribution with these feature spaces in the proximity plot (@fig-satlantic b). Importance variables varied by species (importance heat map @fig-satlantic c), with IPI measurements having relatively low importance (@fig-satlantic c). The distribution of votes (@fig-satlantic d) shows relatively weak classification results; most events consisted of \<50% of the trees voting for the correct species.

@@

The Hawaii dataset consisted of 13 drifting recorders deployed within the Main Hawaiian Islands; species included Blainville's beaked whale (n = 521), Cross Seamount beaked whale (n = 76), Cuvier's beaked whale (n = 201), and Longman's beaked whale (n = 122). BANTER classification models included (1) echolocation pulses (EC), and (2) echolocation pulses and IPI (EC_IPI).

The most accurate model considered the echolocation pulses only (EC Model, Detector Model: *sampsize* = 10, *ntree* = 5,000; Event model: *sampsize* = 4, *ntree* = 10,000) and provided an overall correct classification rate of 92.3% for all four species. Classification scores ranged from a low of 86% (Cuvier's beaked whale) to 97.5% (Longman's beaked whale, @fig-hawaii). All classification results were greater than expected (Prior in @fig-hawaii a). Results from the EC model are presented here; results from EC_IPI model can be found in supplementary materials (Supplement Fig. 4).

Distinctive clusters can be seen for Longman's and Blainville's beaked whales, although there is overlap on these first two dimensions of the clusters for Cuvier's and Cross Seamount beaked whales (@fig-hawaii b). This overlap is resolved with different importance features for Cuvier's and Cross Seamount beaked whales as seen in the importance heat map (@fig-hawaii c), resulting in the low misclassification rates for these species (@fig-hawaii a confusion matrix). The distribution of votes (@fig-hawaii d) show strong classification results for most Blainville's, Cross Seamount, and Longman's beaked whales, with lower classification strength for many Cuvier's beaked whale events.

## EPacific

The EPacific dataset consisted of 15 drifting recorders deployed off the U.S. West Coast, and species included Baird's beaked whale (n = 29), unidentified beaked whale 'BW43' (n = 125), Cross Seamount beaked whale (n = 6), Cuvier's beaked whale (n = 926), Hubbs' beaked whale (n = 66), and Stejneger's beaked whale (n = 42). BANTER classification models included (1) echolocation pulses (EC), (2) echolocation pulses and IPI (EC_IPI), and (3) echolocation pulses, IPI, and environmental features (EC_IPI_ENV).

The best model considered the echolocation pulses, IPI, and environmental features (EC_IPI_ENV Model, Detector Model: *sampsize* = 1, *ntree* = 10,000; Event model: *sampsize* = 5, *ntree* = 10,000) and provided an overall correct classification rate of 91.9% for all six species. Classification scores ranged from a low of 91.2% (BW43) to 100% for Cross Seamount beaked whale (@fig-epacific a). All classification results were greater than expected (Prior in @fig-epacific a). Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs. 5).

There is considerable overlap in the primary feature space for all but BW43, as shown in the proximity plot (@fig-epacific b). IPI was among the most important classification variable for all species (@fig-epacific c importance heat map). The distribution of votes (@fig-epacific d) show relatively strong classification results for Baird's, Hubbs', and Stejneger's beaked whales.
